I/O simulation model with Linux page cache, integration and evaluation in SimGrid framework

### 2020-06-21 paper_io_sim.tex
Section 3 revised

			In our I/O simulation model, we try to simulate the write back page cache 
			mechanisms including memory I/O, cache eviction and data flushing as we 
			believe that they are the most important factors that affect the I/O time.

			Memory is modeled with capacity, read bandwidth and write 
			bandwidth. When data is read from or written to memory, the actual 
			read/write time is calculated with the amount of data and memory 
			read/write bandwidth:

@@

			For periodical flushing, because dirty data can be flushed during CPU time 
			or cache read with disk write bandwidth, the amount of flushed data is 
			calculated with disk bandwidth and flushing duration.

			Similarly, our cache eviction simulation function frees up the page cache 
			by traversing and deleting least recently used clean data blocks 
			in the inactive list.
			The amount of evicted data is passed in and data blocks are deleted 
			from the inactive list until total evicted data reaches the amount 
			passed in, or there is no clean block left in the list.
			If the last evicted block is not entirely evicted, the block is split, 
			and only one block is deleted.
			Our cache eviction simulation does not add up to the simulated time 
			since cache eviction time is negligible in real systems.

			We encapsulate the models of memory and page cache LRU lists in 
			\textit{MemoryManager} class. 
			The functionalities of this class are exposed to the simulation  
			of I/O activities including file read and file write. 
			We create \textit{IOController} class which simulates file read and 
			file write by interacting with the MemoryManager class, 
			exposing APIs to user simulated applications. 
			These two class act as an extra layer with kernel-like functionalities as 
			illustrated in Figure~\ref{fig:interaction}. 
			Simulated applications can simply request for a file read 
			or file write by calling the APIs of this layer instead of 
			calling directly to storage simulation models.
			Because every computer has its own memory and system configurations, 
			one IOController object and one MemoryManager object are created 
			for each simulated machine.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{$r_m$}{memory read bandwidth}
						\Desc{$r_d$}{disk read bandwidth}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
					\State rt = rt + mm.flush(2*fs - mm.cached(fn) - mm.avail\_mem)
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State mm.cache\_read(fn) 
						\State rt = rt + mm.cached(fn) / $r_m$
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State mm.disk\_read(fs - mm.cached(fn), fn)
    					\State rt = rt + (fs - mm.cached(fn)) / $r_d$
					\EndIf

@@

			The algorithm used to simulate file read is described in 
			Algorithm~\ref{alg:read}. 
			In Linux systems with write back storage devices, data can be either read 
			from cache with memory bandwidth or read from disk with disk bandwidth. 
			Therefore, total read time can be considered as a linear function 
			of the read time with each of these bandwidths, 
			enabling us to generalize reading process as two separate phases, 
			cache read and disk read. 
			In our simulator, when a user simulated application reads a file, 
			it sends a file read request to IOController. 
			Then, IOController calculates the amount of required memory to read 
			the file based on the amount of available memory, file size, and 
			the amount of cached data of that file (line 10). 
			If there is not enough available memory, dirty data is flushed to disk 
			to increase the amount of memory available by calling flush function 
			of MemoryManager with the amount of flushed data, flushing time is 
			simulated and added to total run time (line 11-12).
			After data is flushed, MemoryManager is asked to evict data 
			from cache to accommodate the requested file (line 13-14). 
			Next, if file is cached either partially or entirely, MemoryManager is 
			called to re-access and update this cached data (line 16). 
			If all file data is cached, periodical flushing is called is to flush 
			dirty data during the cache read, and time to read from cache is added to 
			total run time, (line 18-20). 
			If file is not cached or partially cached, a disk read is required (line 23) 
			This blocks flushing and makes the cache read time neglectable. 
			Finally, data is read from disk to cache and managed 
			by MemoryManager, disk read time is estimated with disk read 
			bandwidth and added to total run time (line 24-25).

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{$w_m$}{memory write bandwidth}
						\Desc{$w_d$}{disk write bandwidth}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
					\State dirty\_left = $d$ * mm.avail\_mem - mm.dirty
					\If {dirty\_left $>$ 0}
   						\State free\_amt = min(fs, $M$) \Comment{$M$ in Equation
   						~\ref{equa:freeamt}}
    					\State mm.evict(free\_amt - mm.free\_mem)
    					\State free\_time = free\_amt / $w_m$
    					\State mm.write(fn, free\_amt, dirty = free\_amt - free\_time * $w_d$)
   						\State rt = rt + free\_time
    				\EndIf

					\If {free\_amt $\geq$ fs}
						\Return rt
					\EndIf

					\State throttled\_amt = fs - free\_amt \Comment{Written with disk bandwidth}
					\State req = max(0, throttled\_amt  - mm.free\_mem)
					\State evicted\_amt = min(req, mm.evictable)
					\State mm.evict(evicted\_amt)
					\State cached\_amt = min(mm.free\_mem, throttled\_amt)
					\State mm.write(fn, cached = cached\_amt, dirty = 0)
					\State rt = rt + throttled\_amt /$w_d$

@@

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			Similar to the file read simulation, a file can also be written with 
			two different bandwidths, memory bandwidth and disk bandwidth. 
			Files can be written with memory bandwidth before the amount 
			of dirty data reaches dirty\_ratio. Thus, our algorithm initially checks 
			this condition (line 10), and estimates the maximum amount 
			of data can be written with memory bandwidth with the following 
			formula:

			\begin{equation}
				M = max(0, \frac{(d*C - D)*w_m}{w_m - w_d})\label{equa:freeamt}
			\end{equation}

			where:
			\begin{itemize}
				\item $M$ is the maximum amount of data that can be written with memory 
				write bandwidth
				\item $d$ is dirty ratio
				\item $C$ is the total available memory
				\item $D$ is the amount of dirty data
				\item $w_m$ is the memory write bandwidth 
				\item $w_d$ is the disk write bandwidth			
			\end{itemize}

			In this formula, we assume that data is written with with memory 
			bandwidth until dirty\_ratio is reached, the kernel writes data 
			to cache and flushes dirty data to disk concurrently. 
			If file size is smaller than or equal to $M$, the whole file is 
			written to cache as dirty data with memory bandwidth (line 12). 
			Otherwise, the remaining file data is written with disk bandwidth (line 20). 
			Before simulating cache write, MemoryManager is asked to evict 
			the page cache to accommodate written data (line 13). 
			Then, data is written to cache as dirty data with memory 
			bandwidth, and write time is added to total simulated time (line 14-16). 
			Because cache write and data flushing are concurrent in our algorithm, 
			the actual amount of written dirty data reduces by the amount 
			of data during cache write time (line 15). 
			If not all data is written with memory write bandwidth, the remaining data 
			is written with disk write bandwidth. This can simulate both cases that 
			data is written directly to disk, or applications wait for dirty data 
			to be flushed, and evicted to write new data. 
			If the amount of free memory is not enough for the remaining file data, 
			evictable memory is evicted (line 20-23). 
			Here, the simulator does not guarantee that the amount of free memory 
			after eviction can accommodate the remaining data. 
			In the case that free memory is less than remaining data, the amount to be 
			cached is limited by the amount of free memory (line 24). 
			However, the kernel still flushes, evicts and writes new data concurrently, 
			and remaining data is still written to cache with disk write bandwidth.
			Thus, we can simulate the time to write the remaining data with 
			disk bandwidth and ask MemoryManager to add clean cached data (line 25). 
			The simulated time to write width disk bandwidth is added to total 
			simulated run time and returned to simulated applications (line 26).

>>>>

			In our I/O simulation model, we try to simulate the write back page cache 
			mechanisms including memory I/O, cache eviction and data flushing as we 
			believe that they are the most important factors that affect the I/O time. 
			We separate our model into components and encapsulate them in a 
			kernel-like layer described in Figure~\ref{fig:interaction}. 
			This layer consists of two main classes, MemoryManager and IOManager.
			MemoryManager is responsible for memory simulation, which includes 
			memory I/O, flushing, cache eviction, and exposed its functions 
			to IOManager. The IOController class models higher abstraction with the 
			simulation of file read and file write.
			With this layer, simulated applications can simply request for 
			file reads and file writes by calling functions of IOManager class 
			instead of calling directly to storage simulation models.

			\subsubsection{Memory simulation model}

			We model memory with capacity, read bandwidth and write bandwidth. 
			When data is read from or written to memory, the actual 
			read/write time is calculated with the amount of data and memory 
			read/write bandwidth:

@@

			For periodical flushing, our simulator searches in two page cache 
			LRU lists and flushes all the expired dirty data block. 
			A dirty data block is considered expired if the time from its last access 
			to current simulated time is longer than the predefined expire time. 
			Then the flushing time is calculated with total amount of flushed data.

			Similarly to data flushing, our cache eviction simulation function frees up 
			the page cache by traversing and deleting least recently used clean 
			data blocks in the inactive list.
			The amount of evicted data is passed in and data blocks are deleted 
			from the inactive list until total evicted data reaches the amount 
			passed in, or there is no clean block left in the list.
			If the last evicted block is not entirely evicted, the block is split, 
			and only one block is deleted.
			Our cache eviction simulation does not add up to the simulated time 
			since cache eviction time is negligible in real systems.

			\subsubsection{File I/O simulation model}

			Simulated applications can simply request for a file read 
			or file write by calling the APIs of this layer instead of 
			calling directly to storage simulation models.
			Because every computer has its own memory and system configurations, 
			one IOController object and one MemoryManager object are created 
			for each simulated machine.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{rm}{memory read bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
					\State rt = rt + mm.flush(2*fs - mm.cached(fn) - mm.evictable\_mem) 
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State mm.cache\_read(fn)  \Comment{Update blocks in LRU lists}
    					\State pft = mm.periodic\_flush(rt) \Comment{Periodical flushing time}
						\State rt = rt + max(mm.cached(fn) / rm, pft) \Comment{Add the longer}
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State rt = rt + mm.periodic\_flush(rt) \Comment{Another flushing interval}
						\State mm.disk\_read(fs - mm.cached(fn), fn)
    					\State rt = rt + sm.read(fs - mm.cached(fn))
					\EndIf

@@

			The algorithm used to simulate file read is described in 
			Algorithm~\ref{alg:read}. 
			In Linux systems with write back storage devices, data can be either read 
			from cache with memory bandwidth or read from disk with disk bandwidth. 
			Therefore, total read time can be considered as a linear function 
			of the read time with each of these bandwidths, 
			enabling us to generalize reading process as two separate phases, 
			cache read and disk read. 
			In our simulator, when a user simulated application reads a file, 
			it sends a file read request to IOController. 
			First, dirty data is flushed to disk if needed to make the amount of 
			available memory enough to read the file.
			Data flushing function in MemoryManager is called, flushing time is 
			simulated and added to total run time (line 10).
			After data is flushed, MemoryManager is asked to evict data 
			from cache to accommodate the requested file (line 11). 
			Next, if file is cached either partially or entirely, MemoryManager is 
			called to re-access and update this cached data (line 13). 
			Then, periodical flushing is called to flushed expired dirty data 
			from cache (line 14).
			Since periodical flushing and cache read can be concurrent, the longer 
			time is added to simulated run time (line 15).
			If file is not cached or partially cached, a disk read is required (line 17). 
			Periodical flushing is called once again, but this time the flushing time is 
			added to total simulated time since flushing and disk read are time-shared, 
			or sequential in other words (line 18).
			Finally, data is read from disk to cache and managed by MemoryManager, 
			disk read time estimated by the storage model is added to 
			total simulated run time (line 19-20).

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{wm}{memory write bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
   					\State pft = mm.periodic\_flush(rt)
					\State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
					\State mem\_amt = min(fs, remain\_dirty)
					\If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
    					\State mm.evict(mem\_amt - mm.free\_mem)
    					\State mm.write(fn, mem\_amt)
    				\EndIf
    				\State rt = rt + max(mem\_amt / wm, pft) \Comment{Concurrent processes}
					\If {mem\_amt $<$ fs}  \Comment{Write with disk bandwidth}
						\State mm.flushed(fs - mem\_amt)  \Comment{Concurrent flushing}
						\State mm.evict(fs - mem\_amt  - mm.free\_mem) 
						\State mm.write(fn, min(fs - mem\_amt, mm.free\_mem))
						\State rt = rt + sm.write(fs - mem\_amt)
					\EndIf

@@

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			Initially, periodical flushing is called to flush expired dirty data 
			at the beginning of the write (line 10).
			Similar to the file read simulation, a file can also be written with 
			two different bandwidths, memory bandwidth and disk bandwidth. 
			Files can be written with memory bandwidth before the amount 
			of dirty data reaches dirty\_ratio. Thus, our algorithm initially checks 
			the amount of data that can be written with memory write 
			bandwidth (line 11-12).
			If this amount is greater than 0, MemoryManager is asked to evict 
			data from cache if needed and write this amount of data to cache 
			(line 14-15).
			The simulated run time can be added with the longer time of cache write 
			time and periodical flushing time because these can happen 
			concurrently (line 17).
			The amount of data that is not written with memory bandwidth is written 
			with disk bandwidth (line 18). 
			Now, the dirty ratio threshold is reached, we have to flush and evict 
			data from cache as much as possible so that 
			the remaining data can be written to cache (line 19-20). 
			Here, MemoryManager does not guarantee that the amount of flushed 
			and evicted data is enough for the remaining data. In the case it is 
			less than the required amount, the remaining data is written to cache 
			but then can be flushed and evicted right away. Thus, the amount 
			of remaining data that is written and kept in cache after the write 
			is limited by the amount free memory (line 21). 
			Finally, the write time is simulated, added to the total time and 
			returned to simulated applications (line 22).
